---
title: "Fit models"
author: "Max Lindmark"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
  df_print: paged
pdf_document: default
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 12,
  fig.asp = 0.618,
  fig.align ='center'
)
```

# Load libraries

```{r load libraries, message=FALSE, warning=TRUE}
# Load libraries, install if needed
library(tidyverse)
library(tidylog)
library(RCurl)
library(RColorBrewer)
library(sdmTMB)
library(devtools)
library(patchwork)
library(sf)
library(ggeffects)
library(visreg)
library(boot)

# Source utm function
source_url("https://raw.githubusercontent.com/maxlindmark/marine-litter/main/R/functions/lon-lat-utm.R")

# Continuous colors
# options(ggplot2.continuous.colour = "viridis")
# 
# # Discrete colors
# scale_colour_discrete <- function(...) {
#   scale_colour_brewer(palette = "Dark2")
# }
# 
# scale_fill_discrete <- function(...) {
#   scale_fill_brewer(palette = "Dark2")
# }
```

# Make map plots

```{r}
# Normally I'd source code for map plots, but this caused errors when ggplotting when knitting only, not when running the for loop in a fresh session... 
# change to url once we have the final one
#source("/Users/maxlindmark/Dropbox/Max work/R/marine-litter/R/functions/map-plot.R")
# Seem I have to run this to avoid error when knitting (the ggplotting inside the for loop)

# Specify map ranges
ymin = 54; ymax = 60; xmin = 2; xmax = 21

map_data <- rnaturalearth::ne_countries(
  scale = "medium",
  returnclass = "sf",
  continent = "europe")

# Crop the polygon for plotting and efficiency:
# st_bbox(map_data) # find the rough coordinates
sf::sf_use_s2(FALSE)

swe_coast <- suppressWarnings(suppressMessages(
  st_crop(map_data,
          c(xmin = xmin, ymin = ymin, xmax = xmax, ymax = ymax))))

ggplot(swe_coast) + 
  geom_sf()

# Transform our map into UTM 33 coordinates, which is the equal-area projection we fit in:
utm_zone33 <- 32633
swe_coast_proj <- sf::st_transform(swe_coast, crs = utm_zone33)

ggplot(swe_coast_proj) + 
  geom_sf()

# Define plotting theme for main plot
theme_plot <- function(base_size = 11, base_family = "") {
  theme_light(base_size = base_size, base_family = "") +
    theme(
      axis.text = element_text(color = "grey5"),
      legend.position = "bottom",
      legend.key.height = unit(0.2, "cm"),
      legend.margin = margin(0, 0, 0, 0),
      legend.box.margin = margin(-5, -5, -5, -5),
      strip.background = element_rect(fill = "grey95"),
      strip.text = element_text(color = "grey10"),
      strip.text.x = element_text(margin = margin(b = 2, t = 2), color = "grey10", size = 10),
      panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
      panel.background = element_blank()
      #, axis.line = element_line(colour = "grey20"),
    )
}

# Make default base map plot
#sf::st_boundary(swe_coast_proj)
xmin2 <- -61896.44*1.005
xmax2 <- 893074.5*0.91
ymin2 <- 5983578*1.025
ymax2 <- 6691902*0.99

plot_map <-
  ggplot(swe_coast_proj) +
  xlim(xmin2, xmax2) +
  ylim(ymin2, ymax2) +
  labs(x = "Longitude", y = "Latitude") +
  geom_sf(size = 0.3) +
  theme(axis.text.x = element_text(angle = 90)) +
  theme_plot() +
  NULL

plot_map_west <-
  ggplot(swe_coast_proj) +
  xlim(200000, xmax2*0.45) +
  ylim(ymin2*1.015, ymax2*0.99) +
  labs(x = "Longitude", y = "Latitude") +
  geom_sf(size = 0.3) +
  theme_plot() +
  theme(axis.text.x = element_text(angle = 90)) +
  NULL
```

# Read data
Filter litter categories that have enough data to work with (hard to fit models to e.g., glass and metal since they occur so rarely in the data, some years have nothing. Could consider pooling years for those. See exploratory model fitting script (doesn't exist yet))

```{r}
# Read and make data long so that I can for loop through all categories
biom <- read_csv("data/west_coast_litter_biomass.csv") %>% 
  mutate(year_f = as.factor(year),
         quarter_f = as.factor(quarter),
         depth_sc = (depth - mean(depth)) / sd(depth)) %>% 
  rename(plastic = plast) %>% 
  pivot_longer(c(plastic, sup, fishery),
               names_to = "litter_category",
               values_to = "density") %>% 
  filter(litter_category %in% c("fishery", "plastic", "sup"))

num <- read_csv("data/west_coast_litter_numbers.csv") %>% 
  mutate(year_f = as.factor(year),
         quarter_f = as.factor(quarter),
         depth_sc = (depth - mean(depth)) /sd(depth)) %>% 
  rename(plastic = plast) %>% 
  pivot_longer(c(plastic, sup, fishery),
               names_to = "litter_category",
               values_to = "density") %>% 
  mutate(number = density * swept_area_km2) %>% 
  filter(litter_category %in% c("fishery", "plastic", "sup"))

binom <- read_csv("data/west_coast_litter_biomass.csv") %>% 
  mutate(year_f = as.factor(year),
         quarter_f = as.factor(quarter),
         depth_sc = (depth - mean(depth)) / sd(depth)) %>% 
  rename(other = diverse,
         natural = naturmaterial,
         glass = glas,
         metal = metall,
         rubber = gummi) %>% 
  pivot_longer(c(other, natural, glass, metal, rubber),
               names_to = "litter_category",
               values_to = "density") %>% 
  filter(!litter_category %in% c("fishery", "plastic", "sup")) %>% 
  mutate(present = ifelse(density == 0, 0, 1))

# Load pred grid
pred_grid_west <- read_csv("data/pred_grid_west.csv") %>% 
  mutate(year_f = as.factor(year),
         quarter_f = as.factor(1),
         depth_sc = (depth - mean(num$depth)) / sd(num$depth)) %>% 
  mutate(X = X*1000,
         Y = Y*1000)

# hist(pred_grid_west$depth_sc)
# hist(pred_grid_west$depth)
# sd(pred_grid_west$depth)

plot_map_west + 
  theme_plot(base_size = 14) +
  geom_point(data = pred_grid_west, aes(X, Y), alpha = 0.5, size = 0.01) 

ggsave("figures/supp/pred_grid.png", dpi = 300)
```

# Fit models
For plastic, sup and fishery I have enough data to fit spatial models with both biomass density (Tweedie) and numbers (Poisson). For the other categories, I will only fit presence/absence models, because that's almost what we got anyway.

```{r}
# https://haakonbakkagit.github.io/btopic104.html
# https://haakonbakkagit.github.io/btopic114.html

# max.edge <- mean(c(diff(range(num$X)), diff(range(num$Y)))) / 15
# cutoff <- max.edge/5

mesh <- make_mesh(num %>% filter(litter_category == "sup"), c("X", "Y"), cutoff = 4)
#mesh <- make_mesh(num %>% filter(litter_category == "sup"), c("X", "Y"), type = "kmeans", n_knots = 50)
plot(mesh)
```

### Test model to see 

```{r first test to check residual correlation}

    dd <- biom %>%
        filter(litter_category == "fishery") %>%
        droplevels()

    m0 <- sdmTMB(
      data = dd,
      formula = density ~ 0 + year_f + depth_sc,
      mesh = mesh,
      family = tweedie(),
      spatial = "off",
      time = "year",
      spatiotemporal = "off"
      )

    dd$m0_resids <- residuals(m0) # randomized quantile residuals

    hist(dd$m0_resids)
    
    m1 <- sdmTMB(
      data = dd,
      formula = density ~ 0 + year_f + depth_sc,
      mesh = mesh,
      family = tweedie(),
      spatial = "on",
      time = "year",
      spatiotemporal = "IID"
      )

    dd$m1_resids <- residuals(m1) # randomized quantile residuals

    dd2 <- dd %>%
      pivot_longer(c(m0_resids, m1_resids))

    ggplot(dd2 %>% filter(value > -3 & value < 3), aes(X, Y, col = value)) +
      scale_colour_gradient2() +
      geom_point(size = 2) +
      facet_grid(name~year) +
      theme_minimal() +
      coord_fixed() +
      theme(panel.grid.major = element_blank(),
            panel.grid.minor = element_blank(),
            panel.background = element_blank())

    # Check spatial autocorrelation
    pred_fixed <- predict(m0)$est
    s <- simulate(m0, nsim = 500)
    r <- DHARMa::createDHARMa(
      simulatedResponse = s,
      observedResponse = dd$density,
      fittedPredictedResponse = pred_fixed
      )
    plot(r)
    DHARMa::testSpatialAutocorrelation(r, x = dd$X, y = dd$Y)

    # In this case, it doesn't seem like a very clear spatial correlation

    print(m1)
        
```

## Binomial models

```{r}
data_list_coef <- list()
data_list_pred <- list()

# Year as a random effect because in some years we don't have any observations at all of presences

for(i in unique(binom$litter_category)) {
    
    dd <- binom %>%
        filter(litter_category == i) %>%
        droplevels()
    
    m <- sdmTMB(
      data = dd,
      formula = present ~ depth_sc + (1|year_f),
      offset = dd$swept_area_km2, #log
      mesh = mesh,
      family = binomial(link = "logit"),
      spatial = "off",
      time = "year",
      spatiotemporal = "off",
      control = sdmTMBcontrol(newton_loops = 1),
      )
    
    sanity(m)
    tidy(m, conf.int = TRUE)
    
    data_list_coef[[i]] <- tidy(m, conf.int = TRUE) %>% mutate(model = paste("binom", i, sep = "_"))

    # Plot residuals
    mcmc_res <- residuals(m, type = "mle-mcmc", mcmc_iter = 201, mcmc_warmup = 200)
    qqnorm(mcmc_res, asp = 1, main = paste("Normal Q-Q Plot", i, sep = " "))
    qqline(mcmc_res)
    
    # Check for spatial autocorrelation
    pred_fixed <- m$family$linkinv(predict(m)$est)
    s <- simulate(m, nsim = 500)
    r <- DHARMa::createDHARMa(
      simulatedResponse = s,
      observedResponse = dd$present,
      fittedPredictedResponse = pred_fixed
      )
    plot(r)
    DHARMa::testSpatialAutocorrelation(r, x = dd$X, y = dd$Y)
    
    # Plot conditional/marginal effects
    visreg(m, xvar = "year_f", scale = "response")
    
    nd <- data.frame(
      depth_sc = mean(dd$depth_sc),
      year = unique(as.integer(as.character(dd$year_f)))) %>%
      mutate(year_f = as.factor(year),
             X = 0,
             Y = 0)

    p <- predict(m, newdata = nd, se_fit = TRUE, re_form = NULL)
    
    print(ggplot(p, aes(year, inv.logit(est),
                  ymin = inv.logit(est - 1.96 * est_se),
                  ymax = inv.logit(est + 1.96 * est_se)
                  )) +
      geom_line() +
      geom_ribbon(alpha = 0.4) +
      coord_cartesian(expand = F))
    
    data_list_pred[[i]] <- p %>% mutate(model = i)
    
    # Save model object
    saveRDS(m, paste("output/models/binom_", i, ".rds", sep = ""))
    
    print(ggplot(p, aes(year, y = inv.logit(est), ymax = inv.logit(est + 1.96*est_se), ymin = inv.logit(est - 1.96*est_se))) +
      geom_line() +
      geom_line(data = dd %>%
                  group_by(year) %>%
                  summarise(present = mean(present)),
                aes(year, present, color = "Data (mean)"), linetype = 2,
                inherit.aes = FALSE) + # Add data
      geom_ribbon(alpha = 0.2) +
      scale_color_brewer(palette = "Set1", name = "") +
      scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) + 
      theme_plot() +
      xlab('Year') +
      ylab('Mean estimate') +
      NULL)
    
    ggsave(paste("figures/supp/mean_pred_comp_binom_", i, ".png", sep = ""), dpi = 300)
    
}

# Save predictions and sims as data frames
dat_coef_binom <- dplyr::bind_rows(data_list_coef)
dat_pred_binom <- dplyr::bind_rows(data_list_pred)

write_csv(dat_coef_binom, "output/dat_coef_binom.csv")
write_csv(dat_pred_binom, "output/dat_pred_binom.csv")
```

## Poisson/negative binomial models

```{r}
data_list_coef <- list()
data_list_pred <- list()
data_list_sim <- list()
data_list_sims <- list()

for(i in unique(num$litter_category)) {
    
    dd <- num %>%
        filter(litter_category == i) %>%
        droplevels()
    
    m <- sdmTMB(
      data = dd,
      formula = number ~ -1 + year_f + depth_sc,
      mesh = mesh,
      offset = dd$swept_area_km2, #should be log
      family = nbinom2(link = "log"),
      spatial = "on",
      time = "year",
      spatiotemporal = "off",
      control = sdmTMBcontrol(newton_loops = 1)
      )
    
    sanity(m)
    tidy(m, conf.int = TRUE)
    data_list_coef[[i]] <- tidy(m, conf.int = TRUE) %>% mutate(model = paste("poisson", i, sep = "_"))

    # Plot residuals
    mcmc_res <- residuals(m, type = "mle-mcmc", mcmc_iter = 201, mcmc_warmup = 200)
    qqnorm(mcmc_res, asp = 1, main = paste("Normal Q-Q Plot", i, sep = " "))
    qqline(mcmc_res)
    
    # Plot conditional/marginal effects
    visreg(m, xvar = "year_f", scale = "response")

    # Save model object
    saveRDS(m, paste("output/models/numbers_", i, ".rds", sep = ""))
    
    # Predict on grid
    pred <- predict(m, newdata = pred_grid_west) %>% 
      mutate(model = i)
    
    data_list_pred[[i]] <- pred
  
    # Get sims
    ncells <- filter(pred_grid_west, year == max(pred_grid_west$year)) %>% nrow()
    nsim <- 500
    sim <- predict(m, newdata = pred_grid_west, nsim = nsim)
    
    # Plot CV in space
    # Just plot last year
    sim_last <- sim[pred_grid_west$year == max(pred_grid_west$year), ]
    
    pred_last <- pred[pred$year == max(pred_grid_west$year), ]
    pred_last$cv <- round(apply(exp(sim_last), 1, function(x) sd(x) / mean(x)), 2)
    
    print(plot_map_west + 
      geom_raster(data = pred_last, aes(X, Y, fill = cv)) +
      scale_fill_viridis_c() + 
      geom_sf(size = 0.1) + 
      NULL)
    
    ggsave(paste("figures/supp/cv_numbers_", i, ".png", sep = ""))
    
    # Get index & full index (i.e. returning all sims)
    index_sim <- get_index_sims(sim,
                                area = rep(1/ncells, nrow(sim))) %>% mutate(model = i) # Note that we just get the means here
    
    data_list_sim[[i]] <- index_sim
    
    index_sim_full <- get_index_sims(sim,
                                     area = rep(1/ncells, nrow(sim)),
                                     return_sims = TRUE) %>% mutate(model = i) # Note that we just get the means here
    
    data_list_sims[[i]] <- index_sim_full
    
    
    # See how mean index compares to data
    
    print(ggplot(index_sim, aes(year, y = est, ymin = lwr, ymax = upr)) +
      geom_line() +
      geom_line(data = dd %>%
                  group_by(year) %>%
                  summarise(mean_number = mean(number)),
                aes(year, mean_number, color = "Data (mean)"), linetype = 2,
                inherit.aes = FALSE) + # Add data
      geom_ribbon(alpha = 0.2) +
      scale_color_brewer(palette = "Set1", name = "") +
      scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) + 
      theme_plot() +
      xlab('Year') +
      ylab('Mean estimate') +
      NULL)
    
    ggsave(paste("figures/supp/mean_pred_comp_num_", i, ".png", sep = ""), dpi = 300)
    
}

# Save predictions and sims as data frames
dat_coef_num <- dplyr::bind_rows(data_list_coef)
dat_pred_num <- dplyr::bind_rows(data_list_pred)
dat_sim_num <- dplyr::bind_rows(data_list_sim)
dat_sims_num <- dplyr::bind_rows(data_list_sims)

write_csv(dat_coef_num, "output/dat_coef_num.csv")
write_csv(dat_pred_num, "output/dat_pred_num.csv")
write_csv(dat_sim_num, "output/dat_sim_num.csv")
write_csv(dat_sims_num, "output/dat_sims_num.csv")
```

## Fit tweedie biomass density models

```{r}
data_list_coef <- list()
data_list_pred <- list()
data_list_sim <- list()
data_list_sims <- list()

for(i in unique(biom$litter_category)) {
    
    dd <- biom %>%
        filter(litter_category == i) %>%
        droplevels()
    
    m <- sdmTMB(
      data = dd,
      formula = density ~ 0 + year_f + depth_sc,
      mesh = mesh,
      family = tweedie(),
      spatial = "on",
      time = "year",
      spatiotemporal = "off"
      )
    
    sanity(m)
    tidy(m, conf.int = TRUE)
    data_list_coef[[i]] <- tidy(m, conf.int = TRUE) %>% mutate(model = paste("tweedie", i, sep = "_"))

    
    # Plot residuals
    mcmc_res <- residuals(m, type = "mle-mcmc", mcmc_iter = 201, mcmc_warmup = 200)
    qqnorm(mcmc_res, asp = 1, main = paste("Normal Q-Q Plot", i, sep = " "))
    qqline(mcmc_res)
    
    # Save model object
    saveRDS(m, paste("output/models/biomass_", i, ".rds", sep = ""))
    
    # Predict on grid
    pred <- predict(m, newdata = pred_grid_west) %>% 
      mutate(model = i)
    
    data_list_pred[[i]] <- pred
  
    # Get sims
    nsim <- 500
    sim <- predict(m, newdata = pred_grid_west, nsim = nsim)
    
    # Plot CV in space
    # Just plot last year
    sim_last <- sim[pred_grid_west$year == max(pred_grid_west$year), ]
    
    pred_last <- pred[pred$year == max(pred_grid_west$year), ]
    pred_last$cv <- round(apply(exp(sim_last), 1, function(x) sd(x) / mean(x)), 2)
    
    print(plot_map_west + 
      geom_raster(data = pred_last, aes(X, Y, fill = cv)) +
      scale_fill_viridis_c() + 
      geom_sf(size = 0.1) + 
      NULL)
    
    ggsave(paste("figures/supp/cv_biomass_", i, ".png", sep = ""))
    
    # Get index & full index (i.e. returning all sims)
    index_sim <- get_index_sims(sim,
                                area = rep(2*2, nrow(sim))) %>% mutate(model = i)
    
    data_list_sim[[i]] <- index_sim
    
    index_sim_full <- get_index_sims(sim,
                                     area = rep(2*2, nrow(sim)),
                                     return_sims = TRUE) %>% mutate(model = i)
    
    data_list_sims[[i]] <- index_sim_full
    
    
    # See how mean index compares to data
    ncells <- filter(pred_grid_west, year == max(pred_grid_west$year)) %>% nrow()
    
    index_sim_avg <- get_index_sims(sim, area = rep(1/ncells, nrow(sim)))
    
    print(ggplot(index_sim_avg, aes(year, y = est, ymin = lwr, ymax = upr)) +
      geom_line() +
      geom_line(data = dd %>%
                  group_by(year) %>%
                  summarise(mean_density = mean(density)),
                aes(year, mean_density, color = "Data (mean)"), linetype = 2,
                inherit.aes = FALSE) + # Add data
      #geom_ribbon(alpha = 0.2) +
      scale_color_brewer(palette = "Set1", name = "") +
      scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) + 
      theme_plot() +
      xlab('Year') +
      ylab('Mean biomass estimate (kg)') +
      NULL)
    
    ggsave(paste("figures/supp/mean_pred_comp_biomass_", i, ".png", sep = ""), dpi = 300)
    
}

# Save predictions and sims as data frames
dat_coef_biom <- dplyr::bind_rows(data_list_coef)
dat_pred_biom <- dplyr::bind_rows(data_list_pred)
dat_sim_biom <- dplyr::bind_rows(data_list_sim)
dat_sims_biom <- dplyr::bind_rows(data_list_sims)

write_csv(dat_coef_biom, "output/dat_coef_biomass.csv")
write_csv(dat_pred_biom, "output/dat_pred_biomass.csv")
write_csv(dat_sim_biom, "output/dat_sim_biomass.csv")
write_csv(dat_sims_biom, "output/dat_sims_biomass.csv")
```

