---
title: "Fit models"
author: "Max Lindmark"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
  df_print: paged
pdf_document: default
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 12,
  fig.asp = 0.618,
  fig.align ='center'
)
```

# Load libraries

```{r load libraries, message=FALSE, warning=TRUE}
# Load libraries, install if needed
library(tidyverse)
library(tidylog)
library(RCurl)
library(RColorBrewer)
library(sdmTMB) # install.packages("sdmTMB", dependencies = TRUE) should get you INLA as well, else do INLA manually first:
library(INLA) # install.packages("INLA",repos=c(getOption("repos"),INLA="https://inla.r-inla-download.org/R/stable"), dep=TRUE)
library(devtools)
library(patchwork)
library(sf)
library(ggeffects)
library(visreg)
library(boot)

# Source utm function and map plots
source_url("https://raw.githubusercontent.com/maxlindmark/marine-litter/main/R/functions/lon-lat-utm.R")
source_url("https://raw.githubusercontent.com/maxlindmark/marine-litter/main/R/functions/map-plot.R")
```

# Read data
Filter litter categories that have enough data to work with (hard to fit models to e.g., glass and metal since they occur so rarely in the data, some years have nothing. Could consider pooling years for those. See exploratory model fitting script (doesn't exist yet))

```{r}
# Read and make data long so that I can for loop through all categories
west <- readr::read_csv("https://raw.githubusercontent.com/maxlindmark/marine-litter/main/data/west_coast_litter.csv") %>% 
  mutate(year_f = as.factor(year),
         quarter_f = as.factor(quarter),
         depth_sc = (depth - mean(depth)) / sd(depth)) %>% 
  pivot_longer(c("fishery_a", "fishery_b", "plast_a", "plast_b", "sup_a", "sup_b", "tot_a", "tot_b"),
               names_to = "litter_category", values_to = "density") 

east <- readr::read_csv("https://raw.githubusercontent.com/maxlindmark/marine-litter/main/data/east_coast_litter.csv") %>% 
  mutate(year_f = as.factor(year),
         quarter_f = as.factor(quarter),
         depth_sc = (depth - mean(depth)) / sd(depth)) %>% 
  pivot_longer(c("fishery_a", "fishery_b", "plast_a", "plast_b", "sup_a", "sup_b", "tot_a", "tot_b"),
               names_to = "litter_category", values_to = "density")

# Load pred grids
pred_grid_west <- read_csv("data/pred_grid_west.csv") %>% 
  mutate(year_f = as.factor(year),
         quarter_f = as.factor(1),
         depth_sc = (depth - mean(west$depth)) / sd(west$depth)) %>% 
  mutate(X = X*1000,
         Y = Y*1000) %>% 
  drop_na(area)

# ggplot(pred_grid_west, aes(X*1000, Y*1000, color = area)) + 
#   geom_point()
  
pred_grid_east <- read_csv("data/pred_grid_east.csv") %>% 
  mutate(year_f = as.factor(year),
         quarter_f = as.factor(1),
         depth_sc = (depth - mean(east$depth)) / sd(east$depth)) %>% 
  mutate(X = X*1000,
         Y = Y*1000)
```

# Fit models
For plastic, sup and fishery I have enough data to fit spatial models with both biomass density (Tweedie) and numbers (Poisson). For the other categories, I will only fit presence/absence models, because that's almost what we got anyway.

```{r}
# https://haakonbakkagit.github.io/btopic104.html
# https://haakonbakkagit.github.io/btopic114.html

# max.edge <- mean(c(diff(range(num$X)), diff(range(num$Y)))) / 15
# cutoff <- max.edge/5

west_mesh <- make_mesh(west %>% filter(litter_category == "fishery_a"), c("X", "Y"), cutoff = 4)
#mesh <- make_mesh(num %>% filter(litter_category == "sup"), c("X", "Y"), type = "kmeans", n_knots = 50)
plot(west_mesh)

east_mesh <- make_mesh(east %>% filter(litter_category == "fishery_a"), c("X", "Y"), cutoff = 4)
#mesh <- make_mesh(num %>% filter(litter_category == "sup"), c("X", "Y"), type = "kmeans", n_knots = 50)
plot(east_mesh)
```

## Fit tweedie biomass density models
### East coast

```{r}
data_list_coef <- list()
data_list_pred <- list()
data_list_sim <- list()
data_list_sims <- list()

for(i in unique(east$litter_category)) {
    
    dd <- east %>%
        filter(litter_category == i) %>%
        droplevels()
    
    m <- sdmTMB(
      data = dd,
      formula = density ~ 0 + year_f + depth_sc,
      mesh = east_mesh,
      family = tweedie(),
      spatial = "on",
      time = "year",
      spatiotemporal = "off"
      )
    
    sanity(m)
    tidy(m, conf.int = TRUE)
    data_list_coef[[i]] <- tidy(m, conf.int = TRUE) %>% mutate(model = paste("east", i, sep = "_"))
    
    # Plot residuals
    mcmc_res <- residuals(m, type = "mle-mcmc", mcmc_iter = 201, mcmc_warmup = 200)
    qqnorm(mcmc_res, asp = 1, main = paste("Normal Q-Q Plot", i, sep = " "))
    qqline(mcmc_res)
    
    # Predict on grid
    pred <- predict(m, newdata = pred_grid_east) %>% 
      mutate(model = i)
    
    data_list_pred[[i]] <- pred
  
    # Get sims
    nsim <- 500
    sim <- predict(m, newdata = pred_grid_east, nsim = nsim)
    
    # Get index & full index (i.e. returning all sims)
    index_sim <- get_index_sims(sim,
                                area = rep(2*2, nrow(sim))) %>% mutate(model = paste("east", i, sep = "_"))
    
    data_list_sim[[i]] <- index_sim
    
    index_sim_full <- get_index_sims(sim,
                                     area = rep(2*2, nrow(sim)),
                                     return_sims = TRUE) %>% mutate(model = paste("east", i, sep = "_"))
    
    data_list_sims[[i]] <- index_sim_full
    
    
    # See how mean index compares to data
    ncells <- filter(pred_grid_east, year == max(pred_grid_east$year)) %>% nrow()
    
    index_sim_avg <- get_index_sims(sim, area = rep(1/ncells, nrow(sim)))
    
    print(ggplot(index_sim_avg, aes(year, y = est, ymin = lwr, ymax = upr)) +
      geom_line() +
      geom_line(data = dd %>%
                  group_by(year) %>%
                  summarise(mean_density = mean(density)),
                aes(year, mean_density, color = "Data (mean)"), linetype = 2,
                inherit.aes = FALSE) + # Add data
      geom_ribbon(alpha = 0.2) +
      scale_color_brewer(palette = "Set1", name = "") +
      scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) + 
      theme_plot() +
      xlab('Year') +
      ylab('Mean density') +
      NULL)
    
    ggsave(paste("figures/supp/east_mean_pred_comp_", i, ".png", sep = ""), dpi = 300)
    
}

# Save predictions and sims as data frames
dat_coef <- dplyr::bind_rows(data_list_coef)
dat_pred <- dplyr::bind_rows(data_list_pred)
dat_sim <- dplyr::bind_rows(data_list_sim)
dat_sims <- dplyr::bind_rows(data_list_sims)

write_csv(dat_coef, "output/east_dat_coef.csv")
write_csv(dat_pred, "output/east_dat_pred.csv")
write_csv(dat_sim, "output/east_dat_sim.csv")
write_csv(dat_sims, "output/east_dat_sims.csv")
```

### West coast

```{r}
data_list_coef <- list()
data_list_pred <- list()
data_list_sim <- list()
data_list_sims <- list()

for(i in unique(west$litter_category)) {
    
    dd <- west %>%
        filter(litter_category == i) %>%
        droplevels()
    
    m <- sdmTMB(
      data = dd,
      formula = density ~ 0 + year_f + depth_sc,
      mesh = west_mesh,
      family = tweedie(),
      spatial = "on",
      time = "year",
      spatiotemporal = "off"
      )
    
    sanity(m)
    tidy(m, conf.int = TRUE)
    data_list_coef[[i]] <- tidy(m, conf.int = TRUE) %>% mutate(model = paste("west", i, sep = "_"))
    
    # Plot residuals
    mcmc_res <- residuals(m, type = "mle-mcmc", mcmc_iter = 201, mcmc_warmup = 200)
    qqnorm(mcmc_res, asp = 1, main = paste("Normal Q-Q Plot", i, sep = " "))
    qqline(mcmc_res)
    
    
    # Predict on grid
    pred <- predict(m, newdata = pred_grid_west) %>% 
      mutate(model = i)
    
    data_list_pred[[i]] <- pred
  
    # Get sims
    nsim <- 500
    sim <- predict(m, newdata = pred_grid_west, nsim = nsim)
    
    # Get index & full index (i.e. returning all sims)
    index_sim <- get_index_sims(sim,
                                area = rep(2*2, nrow(sim))) %>% mutate(model = paste("west", i, sep = "_"))
    
    data_list_sim[[i]] <- index_sim
    
    index_sim_full <- get_index_sims(sim,
                                     area = rep(2*2, nrow(sim)),
                                     return_sims = TRUE) %>% mutate(model = paste("west", i, sep = "_"))
    
    data_list_sims[[i]] <- index_sim_full
    
    
    # See how mean index compares to data
    ncells <- filter(pred_grid_west, year == max(pred_grid_west$year)) %>% nrow()
    
    index_sim_avg <- get_index_sims(sim, area = rep(1/ncells, nrow(sim)))
    
    print(ggplot(index_sim_avg, aes(year, y = est, ymin = lwr, ymax = upr)) +
      geom_line() +
      geom_line(data = dd %>%
                  group_by(year) %>%
                  summarise(mean_density = mean(density)),
                aes(year, mean_density, color = "Data (mean)"), linetype = 2,
                inherit.aes = FALSE) + # Add data
      scale_color_brewer(palette = "Set1", name = "") +
      scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) + 
      theme_plot() +
      xlab('Year') +
      ylab('Mean density') +
      NULL)
    
    ggsave(paste("figures/supp/west_mean_pred_comp_", i, ".png", sep = ""), dpi = 300)
    
}

# Save predictions and sims as data frames
dat_coef <- dplyr::bind_rows(data_list_coef)
dat_pred <- dplyr::bind_rows(data_list_pred)
dat_sim <- dplyr::bind_rows(data_list_sim)
dat_sims <- dplyr::bind_rows(data_list_sims)

write_csv(dat_coef, "output/west_dat_coef.csv")
write_csv(dat_pred, "output/west_dat_pred.csv")
write_csv(dat_sim, "output/west_dat_sim.csv")
write_csv(dat_sims, "output/west_dat_sims.csv")
```

## Fit tweedie biomass density models with year as continious and interaction with time period

### East coast

```{r}
data_list_pred <- list()
data_list_coef <- list()

for(i in unique(east$litter_category)) {
    
    dd <- east %>%
      filter(litter_category == i) %>%
      mutate(time_period = ifelse(year < 2016, "1", "2"),
             time_period = as.factor(time_period),
             year_ct = year - 2013) %>% 
      droplevels()
    
    # Whole time period  
    m1 <- sdmTMB(
      data = dd,
      formula = density ~ year_ct + depth_sc,
      mesh = east_mesh,
      family = tweedie(),
      spatial = "on",
      spatiotemporal = "off"
      )
    
    # Interaction
    m2 <- sdmTMB(
      data = dd,
      formula = density ~ year_ct*time_period + depth_sc,
      mesh = east_mesh,
      family = tweedie(),
      spatial = "on",
      spatiotemporal = "off"
      )
    
    sanity(m1)
    sanity(m2)
    
    tidy(m1)
    tidy(m2)
    
    temp_m1 <- tidy(m1, conf.int = TRUE) 
    temp_m2 <- tidy(m2, conf.int = TRUE) 
    
    nd <- data.frame(year = 2013:2021) %>%
      mutate(depth_sc = mean(dd$depth_sc),
             year_ct = year - 2013,
             time_period = ifelse(year < 2016, "1", "2"),
             time_period = as.factor(time_period))

    m1_preds <- predict(m1, newdata = nd, re_form = NA, re_form_iid = NA)
    m2_preds <- predict(m2, newdata = nd, re_form = NA, re_form_iid = NA)
    
    nd$pred_m1 <- m1_preds$est
    nd$pred_m2 <- m2_preds$est
    
    nd2 <- nd %>% pivot_longer(c(pred_m1, pred_m2))
    
    # Add sig or not
    nd2 <- nd2 %>% 
      mutate(sig_slope = "No",
             sig_slope = ifelse(name == "pred_m1" & temp_m1$conf.low[2] > 0 & temp_m1$conf.high[2] > 0, "Yes", sig_slope),
             sig_slope = ifelse(name == "pred_m1" & temp_m1$conf.low[2] < 0 & temp_m1$conf.high[2] < 0, "Yes", sig_slope)) %>% 
      
      mutate(sig_slope = ifelse(name == "pred_m2" & year < 2016 & temp_m2$conf.low[2] > 0 & temp_m2$conf.high[2] > 0, "Yes", sig_slope),
             sig_slope = ifelse(name == "pred_m2" & year < 2016 & temp_m2$conf.low[2] < 0 & temp_m2$conf.high[2] < 0, "Yes", sig_slope),
             
             sig_slope = ifelse(name == "pred_m2" & year > 2015 & temp_m2$conf.low[2] + temp_m2$conf.low[5] > 0 &
                                  temp_m2$conf.high[2] + temp_m2$conf.low[5] > 0, "Yes", sig_slope),
             sig_slope = ifelse(name == "pred_m2" & year > 2015 & temp_m2$conf.low[2] + temp_m2$conf.low[5] < 0 &
                                  temp_m2$conf.high[2] + temp_m2$conf.low[5] < 0, "Yes", sig_slope))
    
    # ggplot() + 
    #   geom_point(data = dd, aes(year, density)) + 
    #   geom_line(data = nd2, aes(year, exp(value), color = name, linetype = sig_slope)) + 
    #   scale_linetype_manual(values = c(2, 1)) + 
    #   coord_cartesian(ylim = c(0, 10))
    
    nd2 <- nd2 %>% mutate(model = paste("east", i, sep = "_"))
    
    data_list_pred[[i]] <- nd2
    
    # Significant interaction?
    int <- filter(temp_m2, term == "year_ct:time_period2")
    inter <- data.frame(model = paste("east", i, sep = "_")) %>%
      mutate(sig = "No",
             sig = ifelse(int$conf.low > 0 & int$conf.high > 0, "Yes", sig),
             sig = ifelse(int$conf.low < 0 & int$conf.high < 0, "Yes", sig))

    data_list_coef[[i]] <- inter
    
    # Plot residuals
    # mcmc_res <- residuals(m, type = "mle-mcmc", mcmc_iter = 201, mcmc_warmup = 200)
    # qqnorm(mcmc_res, asp = 1, main = paste("Normal Q-Q Plot", i, sep = " "))
    # qqline(mcmc_res)
    
}

# Save predictions
dat_pred <- dplyr::bind_rows(data_list_pred)
dat_coef <- dplyr::bind_rows(data_list_coef)

write_csv(dat_pred, "output/yr_cont_east_dat_pred.csv")
write_csv(dat_coef, "output/yr_cont_inter_east_dat_pred.csv")
```

### West coast

```{r}
data_list_pred <- list()
data_list_coef <- list()

for(i in unique(west$litter_category)) {
    
    dd <- west %>%
      filter(litter_category == i) %>%
      mutate(time_period = ifelse(year < 2016, "1", "2"),
             time_period = as.factor(time_period),
             year_ct = year - 2013) %>% 
      droplevels()
    
    # Whole time period  
    m1 <- sdmTMB(
      data = dd,
      formula = density ~ year_ct + depth_sc,
      mesh = west_mesh,
      family = tweedie(),
      spatial = "on",
      spatiotemporal = "off"
      )
    
    # Interaction
    m2 <- sdmTMB(
      data = dd,
      formula = density ~ year_ct*time_period + depth_sc,
      mesh = west_mesh,
      family = tweedie(),
      spatial = "on",
      spatiotemporal = "off"
      )
    
    sanity(m1)
    sanity(m2)
    
    tidy(m1)
    tidy(m2)
    
    temp_m1 <- tidy(m1, conf.int = TRUE) 
    temp_m2 <- tidy(m2, conf.int = TRUE) 
    
    nd <- data.frame(year = 2013:2021) %>%
      mutate(depth_sc = mean(dd$depth_sc),
             year_ct = year - 2013,
             time_period = ifelse(year < 2016, "1", "2"),
             time_period = as.factor(time_period))

    m1_preds <- predict(m1, newdata = nd, re_form = NA, re_form_iid = NA)
    m2_preds <- predict(m2, newdata = nd, re_form = NA, re_form_iid = NA)
    
    nd$pred_m1 <- m1_preds$est
    nd$pred_m2 <- m2_preds$est
    
    nd2 <- nd %>% pivot_longer(c(pred_m1, pred_m2))
    
    # Add sig or not
    nd2 <- nd2 %>% 
      mutate(sig_slope = "No",
             sig_slope = ifelse(name == "pred_m1" & temp_m1$conf.low[2] > 0 & temp_m1$conf.high[2] > 0, "Yes", sig_slope),
             sig_slope = ifelse(name == "pred_m1" & temp_m1$conf.low[2] < 0 & temp_m1$conf.high[2] < 0, "Yes", sig_slope)) %>% 
      
      mutate(sig_slope = ifelse(name == "pred_m2" & year < 2016 & temp_m2$conf.low[2] > 0 & temp_m2$conf.high[2] > 0, "Yes", sig_slope),
             sig_slope = ifelse(name == "pred_m2" & year < 2016 & temp_m2$conf.low[2] < 0 & temp_m2$conf.high[2] < 0, "Yes", sig_slope),
             
             sig_slope = ifelse(name == "pred_m2" & year > 2015 & temp_m2$conf.low[2] + temp_m2$conf.low[5] > 0 &
                                  temp_m2$conf.high[2] + temp_m2$conf.low[5] > 0, "Yes", sig_slope),
             sig_slope = ifelse(name == "pred_m2" & year > 2015 & temp_m2$conf.low[2] + temp_m2$conf.low[5] < 0 &
                                  temp_m2$conf.high[2] + temp_m2$conf.low[5] < 0, "Yes", sig_slope))
    
    
    nd2 <- nd2 %>% mutate(model = paste("west", i, sep = "_"))
    
    data_list_pred[[i]] <- nd2
    
    # Significant interaction?
    int <- filter(temp_m2, term == "year_ct:time_period2")
    inter <- data.frame(model = paste("west", i, sep = "_")) %>%
      mutate(sig = "No",
             sig = ifelse(int$conf.low > 0 & int$conf.high > 0, "Yes", sig),
             sig = ifelse(int$conf.low < 0 & int$conf.high < 0, "Yes", sig))

    data_list_coef[[i]] <- inter
    
    # Plot residuals
    # mcmc_res <- residuals(m, type = "mle-mcmc", mcmc_iter = 201, mcmc_warmup = 200)
    # qqnorm(mcmc_res, asp = 1, main = paste("Normal Q-Q Plot", i, sep = " "))
    # qqline(mcmc_res)
    
}

# Save predictions
dat_pred <- dplyr::bind_rows(data_list_pred)
dat_coef <- dplyr::bind_rows(data_list_coef)

write_csv(dat_pred, "output/yr_cont_west_dat_pred.csv")
write_csv(dat_coef, "output/yr_cont_inter_west_dat_pred.csv")
```