---
title: "Fit models"
author: "Max Lindmark"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
  df_print: paged
pdf_document: default
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 12,
  fig.asp = 0.618,
  fig.align ='center'
)
```

# Load libraries

```{r load libraries, message=FALSE, warning=TRUE}
# Load libraries, install if needed
library(tidyverse)
library(tidylog)
library(RCurl)
library(RColorBrewer)
library(sdmTMB) # install.packages("sdmTMB", dependencies = TRUE) should get you INLA as well, else do INLA manually first:
library(INLA) # install.packages("INLA",repos=c(getOption("repos"),INLA="https://inla.r-inla-download.org/R/stable"), dep=TRUE)
library(devtools)
library(patchwork)
library(sf)
library(ggeffects)
library(visreg)
library(boot)

# Source utm function and map plots
source_url("https://raw.githubusercontent.com/maxlindmark/marine-litter/main/R/functions/map-plot.R")
```

# Read data
Filter litter categories that have enough data to work with, hard to fit models to e.g., glass and metal since they occur so rarely in the data, some years have nothing. Could consider pooling years for those. See exploratory model fitting script (doesn't exist yet))

```{r}
# Read and make data long so that I can for loop through all categories
west <- readr::read_csv("https://raw.githubusercontent.com/maxlindmark/marine-litter/main/data/west_coast_litter.csv") |> 
  mutate(year_f = as.factor(year),
         quarter_f = as.factor(quarter),
         depth_sc = (depth - mean(depth)) / sd(depth)) |> 
  pivot_longer(c("fishery_a", "fishery_b", "plast_a", "plast_b", "sup_a", "sup_b", "tot_a", "tot_b"),
               names_to = "litter_category", values_to = "density") |> 
  filter(year >= 2013)

east <- readr::read_csv("https://raw.githubusercontent.com/maxlindmark/marine-litter/main/data/east_coast_litter.csv") |> 
  mutate(year_f = as.factor(year),
         quarter_f = as.factor(quarter),
         depth_sc = (depth - mean(depth)) / sd(depth)) |> 
  pivot_longer(c("fishery_a", "fishery_b", "plast_a", "plast_b", "sup_a", "sup_b", "tot_a", "tot_b"),
               names_to = "litter_category", values_to = "density") |> 
  filter(year >= 2013)

# Load pred grids
pred_grid_west <- read_csv("data/pred_grid_west.csv") |> 
  mutate(year_f = as.factor(year),
         quarter_f = as.factor(1),
         depth_sc = (depth - mean(west$depth)) / sd(west$depth)) |> 
  mutate(X = X,
         Y = Y) |> 
  filter(year >= 2013)

pred_grid_east <- read_csv("data/pred_grid_east.csv") |> 
  mutate(year_f = as.factor(year),
         quarter_f = as.factor(1),
         depth_sc = (depth - mean(east$depth)) / sd(east$depth)) |> 
  mutate(X = X,
         Y = Y) |> 
  filter(year >= 2013)
```

# Fit models
For plastic, sup and fishery I have enough data to fit spatial models with both biomass density (Tweedie) and numbers (Poisson).

```{r}
# https://haakonbakkagit.github.io/btopic104.html
# https://haakonbakkagit.github.io/btopic114.html

west_mesh <- make_mesh(west |> filter(litter_category == "fishery_a"), c("X", "Y"), cutoff = 4)
plot(west_mesh)

east_mesh <- make_mesh(east |> filter(litter_category == "fishery_a"), c("X", "Y"),
                       cutoff = 6)
plot(east_mesh)
```

## Fit tweedie models to biomass and abundance density
### East coast

```{r}
data_list_sim <- list()

for(i in unique(east$litter_category)) {
    
    dd <- east |>
        filter(litter_category == i) |>
        droplevels()
    
    m <- sdmTMB(
      data = dd,
      formula = density ~ 0 + year_f + depth_sc,
      mesh = east_mesh,
      family = tweedie(),
      spatial = "on",
      time = "year",
      spatiotemporal = "iid"
      )
    
    sanity(m)
    tidy(m, conf.int = TRUE)

    # Plot residuals
    print(
      simulate(m, nsim = 500, type = "mle-mvn") |>
      dharma_residuals(m, plot = FALSE) |> 
      ggplot(aes(observed, expected)) +
      geom_point(color = "grey30", shape = 21, size = 0.5) +
      geom_abline(col = "tomato3", linewidth = 0.6) +
      theme(aspect.ratio = 1) +
      labs(x = "Observed", y = "Expected")
      )
    
    # Get index
    nsim <- 500
    sim <- predict(m, newdata = pred_grid_east, nsim = nsim)
    index_sim <- get_index_sims(sim,
                                area = rep(2*2, nrow(sim))) |> mutate(model = paste("east", i, sep = "_"))
    
    data_list_sim[[i]] <- index_sim
     
}

# Save predictions and sims as data frames
dat_sim <- dplyr::bind_rows(data_list_sim)

write_csv(dat_sim, "output/east_dat_sim.csv")
```

### West coast

```{r}
data_list_sim <- list()

for(i in unique(west$litter_category)) {
    
    dd <- west |>
        filter(litter_category == i) |>
        droplevels()
    
    m <- sdmTMB(
      data = dd,
      formula = density ~ 0 + year_f + depth_sc,
      mesh = west_mesh,
      family = tweedie(),
      spatial = "on",
      time = "year",
      spatiotemporal = "off",
      )
    
    if (tidy(m, conf.int = TRUE, effects = "ran_pars") |> filter(term == "sigma_O") |> pull(estimate) > 100) {
      
          m <- sdmTMB(
            data = dd,
            formula = density ~ 0 + year_f + depth_sc,
            mesh = west_mesh,
            family = tweedie(),
            spatial = "off",
            time = "year",
            spatiotemporal = "iid",
          ) } else{
        
            m <- m
            
      }
    
    sanity(m)
    
    # Plot residuals
    simulate(m, nsim = 500, type = "mle-mvn") |>
      dharma_residuals(m, plot = FALSE) |> 
      ggplot(aes(observed, expected)) +
      geom_point(color = "grey30", shape = 21, size = 0.5) +
      geom_abline(col = "tomato3", linewidth = 0.6) +
      theme(aspect.ratio = 1) +
      labs(x = "Observed", y = "Expected")
    
    # Get index
    nsim <- 500
    sim <- predict(m, newdata = pred_grid_west, nsim = nsim)
    index_sim <- get_index_sims(sim,
                                area = rep(2*2, nrow(sim))) |> mutate(model = paste("west", i, sep = "_"))
    
    data_list_sim[[i]] <- index_sim
    
}

# Save predictions and sims as data frames
dat_sim <- dplyr::bind_rows(data_list_sim)

write_csv(dat_sim, "output/west_dat_sim.csv")
```
